### 2024-03-29 parts/appendix.tex
rewrite intro
appendix + proofrefs

    The function $f$ is \kl{commutative} if and only if for all $u,v \in
    \Sigma^+$, ${}^t v_1 (M_u M_v  - M_v M_u) v_2 = 0$. The above is a
    polynomial equation involving the coefficients of $M_u$ and $M_v$.
    Let us compute $\overline{S \times S}$ the Zariski closure of $S \times S$.
    The above polynomial vanishes over $S \times S$ if and only
    if it belongs to the Zariski closure, which is decidable
    \cite{HROUPOWO18}.
\end{proof}

\begin{proof}[Proof of \cref{negative-not-nrat:ex}]
    The function $w \mapsto |w|$ is a \kl{$\Nat$-polyregular function}.
    Thus, 
    $P(X) \defined X$ is
    a \kl{$\Nat$-rational polynomial}. Similarly,
    $w \mapsto |w|^2 + 3$ is a \kl{$\Nat$-polyregular function},
    showing that $Q(X) \defined X^2 + 3$
    is a \kl{$\Nat$-rational polynomial}.
    Finally, 
    $T(X) \defined - X$ cannot be 
    a \kl{$\Nat$-rational polynomial} as \kl{$\Nat$-polyregular functions}
    are non-negative.
\end{proof}

@@

\begin{proof}[Proof of
    \cref{n-poly-n-poly:example}]
    The polynomials of $\Nat[\vec{X}]$
    are obtained from basic functions (constant in $\Nat$,
    $X_i$ for some $1 \leq i \leq n$)
    by products and sums. Because the basic functions are
    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions,
    that are closed under these operations, we conclude.
\end{proof}

\begin{proof}[Proof of
    \cref{n-poly-combinatorics:lem}
    ]

    Let $\omega$ be an idempotent power for the finite monoid $M$,
    and
    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
    $f = \pi^\dagger$. That is, 
    $f(w)$ is the sum over all factorizations of $w$
    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.

@@

    Notice that to produce all elements in a given equivalence class
    means adding the factors $u_i^\omega$ where $1 \leq i \leq p$
    back to the normalized version
    of the equivalence class. For all elements in an equivalence class,
    one has to add the same amount of $u_i^\omega$ for $1 \leq i \leq p$,
    and therefore the number of possible choices
    is either constant, or 
    is a product of binomials of the form $\binom{X - l}{s}$
    which has the desired shape.
\end{proof}

\begin{proof}[Proof of \cref{n-rat-correct:lem}]
    Let $f$ be the \kl{commutative}
    \kl{$\Nat$-rational series}
    with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
    it is in fact a \kl{$\Nat$-polyregular function}
    (\cref{polynomial-rational-polyreg:fact}).
    Using \cref{n-poly-combinatorics:lem},
    there exists a number $\omega \in \Nat$
    and polynomial $Q$ with \kl{non-negative}
    \kl{maximal monomials} such that
    for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \prod_{i = 1}^p a_i^{\omega X_i}
        \right)
        = Q(X_1, \dots, X_p)
        = P(\omega X_1, \dots, \omega X_p) 
        \quad .
    \end{equation*}
    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
    as polynomials,
    and in particular that
    the \kl{maximal monomials} of 
    $P$ are \kl{non-negative}.

    Remark that for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
    the polynomial $\restr{P}{\nu}$ continues to be represented
    by a \kl{$\Nat$-polyregular function}, namely
    $f$ partially applied to a word. As a consequence,
    the \kl{maximal monomials} of
    $\restr{P}{\nu}$ are also \kl{non-negative}, 
    and
    we have proven that $P \in \CorrectPoly$.
\end{proof}

@@

\begin{proof}[Proof of 
    \cref{derivation-monomials:fact}
    ]
    Let $M \defined X_1^{k_1} \cdots X_n^{k_n}$,
    where $k_1, \dots, k_n > 0$ and $n > 0$.
    Let us compute
    \begin{align*}
        \Diff{K}{M} &= \prod_{i = 1}^n X_i^{k_i} - \prod_{i = 1}^n (X_i - K)^{k_i} \\
                    &= \sum_{p = 1}^n \frac{K}{X_p - K} \prod_{i = 1}^n (X_i - K)^{k_i} 
    \end{align*}
    Therefore, 
    the \kl{maximal monomials} of $\Diff{K}{M}$
    are precisely
    of the form $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$ for some $1 \leq p \leq i$,
    which \kl{strictly divides} $M$.
    Let $T$ be a \kl{monomial} that \kl{strictly divides} $M$,
    then, $T = \prod_{i = 1}^n X_i^{t_i}$, 
    for all $1 \leq i \leq n$, $t_i \leq k_i$,
    and there exists $1 \leq p \leq n$ such that$ t_j < k_j$.
    In particular, $T$ \kl{divides}
    $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$.
\end{proof}

\begin{proof}[Proof of \cref{derivation-simplifies:lemma}]
    Let $P \in \Rel[\vec{X}]$ and
    $K \in \Nat$.
    Leveraging 
    \cref{derivation-monomials:fact,discrete-deriv-linear:fact,maxi-monomials-submodular:fact}:
    \begin{align*}
        \MaximalMonomials\left(\Diff{K}{P}\right)
        &=
        \MaximalMonomials\left(\sum_{M \in \Monomials(P)} \Diff{K}{M}\right) \\
        &\hoareleq
        \bigcup_{M \in \Monomials(P)} 
        \MaximalMonomials(\Diff{K}{M}) \\
        &\hoarele
        \MaximalMonomials(P)
        \quad .
        \qedhere
    \end{align*}
\end{proof}

\begin{proof}[Proof of \cref{derivation-stabilises-correct:lem}]
    If $P$ is constant, then $P_2 = 0$, and $\Diff{K}{P_1} = 0$,
    hence for $K = 0$, $Q \in \CorrectPoly$.

@@

    Now, remark that the coefficient $\beta$ is obtained as follows:
    \begin{align*}
        \beta &= \sum_{M \in \mathcal{D}_1} \restr{M / T_1}{\nu}
               + \sum_{M \in \mathcal{D}_2} \restr{M / T_1}{\nu} \\
              &\geq \sum_{M \in \mathcal{D}_1} \restr{M / T_1}{\nu}
              - \sum_{M \in \mathcal{D}_2} \left|\restr{M / T_1}{\nu}\right|
              \\
              &\geq
              \sum_{M \in \mathcal{D}_1}
              \left[
              \restr{M / T_1}{\nu}
              -
              \sum_{S \in \mathcal{D}_2^M}
              \left|\restr{S/T_1}{\nu}\right|
              \right]
              \\
              &\geq
              \sum_{M \in \mathcal{D}_1}
              \restr{\mathsf{unit}(M)/T_1}{\nu}
              \left[
              K
              - \alpha D
              \right]
              \\
              &\geq 0 \quad . \qedhere
    \end{align*}
\end{proof}

\begin{proof}[Proof of \cref{derivation-translation:lem}]

@@

    For the effective decision procedure,
    given $P \in \Rel[\vec{X}]$,
    one applies \cref{derivation-translation:lem}
    to obtain a bound $K$, and then 
    computes $\translate{K}(\restr{P}{\nu})$ for all 
    partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots, K}$.
    Then $P \in \CorrectPoly$ if and only if all those polynomials
    belong to $\Nat[\vec{X}]$, which is decidable.
\end{proof}

>>>>

    The function $f$ is \kl{commutative} if and only if for all $u,v \in
    \Sigma^+$, ${}^t v_1 (M_u M_v  - M_v M_u) v_2 = 0$. The above is a
    polynomial equation involving the coefficients of $M_u$ and $M_v$.
    Let us compute $\overline{S \times S}$ the Zariski closure of $S \times S$.
    The above polynomial vanishes over $S \times S$ if and only
    if it belongs to the Zariski closure, which is decidable
    \cite{HOPW18}.
\end{proof}

\begin{proof}[A simpler proof of \cref{decidable-commutative-rat:lemma} in the general case]
    Remark that the group of permutations of $\set{1, \dots, n}$ is generated by
    the cycle $c \defined (n,1, \dots, n-1)$ and the transposition $t \defined (1, 2)$.
    As a consequence, a function $f$ is commutative if and only if
    $f \circ c = f = f \circ t$.
    When $f$ is a rational series,
    $f \circ c$ and $f \circ t$ are both rational series that can be
    effectively computed from $f$, and since equivalence
    of rational series is decidable 
    \cite{BERE10}
    we have obtained a decision procedure.
\end{proof}

\begin{proofof}[negative-not-nrat:ex]
    The function $w \mapsto |w|$ is a \kl{$\Nat$-polyregular function}.
    Thus, 
    $P(X) \defined X$ is
    a \kl{$\Nat$-rational polynomial}. Similarly,
    $w \mapsto |w|^2 + 3$ is a \kl{$\Nat$-polyregular function},
    showing that $Q(X) \defined X^2 + 3$
    is a \kl{$\Nat$-rational polynomial}.
    Finally, 
    $T(X) \defined - X$ cannot be 
    a \kl{$\Nat$-rational polynomial} as \kl{$\Nat$-polyregular functions}
    are non-negative.
\end{proofof}

@@

\begin{proofof}[n-poly-n-poly:example]
    The polynomials of $\Nat[\vec{X}]$
    are obtained from basic functions (constant in $\Nat$,
    $X_i$ for some $1 \leq i \leq n$)
    by products and sums. Because the basic functions are
    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions,
    that are closed under these operations, we conclude.
\end{proofof}

\begin{proofof}[n-poly-combinatorics:lem]
    Let $\omega$ be an idempotent power for the finite monoid $M$,
    and
    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
    $f = \pi^\dagger$. That is, 
    $f(w)$ is the sum over all factorizations of $w$
    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.

@@

    Notice that to produce all elements in a given equivalence class
    means adding the factors $u_i^\omega$ where $1 \leq i \leq p$
    back to the normalized version
    of the equivalence class. For all elements in an equivalence class,
    one has to add the same amount of $u_i^\omega$ for $1 \leq i \leq p$,
    and therefore the number of possible choices
    is either constant, or 
    is a product of binomials of the form $\binom{X - l}{s}$
    which has the desired shape.
\end{proofof}

\begin{proofof}[n-rat-correct:lem]
    Let $f$ be the \kl{commutative}
    \kl{$\Nat$-rational series}
    with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
    it is in fact a \kl{$\Nat$-polyregular function}
    (\cref{polynomial-rational-polyreg:fact}).
    Using \cref{n-poly-combinatorics:lem},
    there exists a number $\omega \in \Nat$
    and polynomial $Q$ with \kl{non-negative}
    \kl{maximal monomials} such that
    for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \prod_{i = 1}^p a_i^{\omega X_i}
        \right)
        = Q(X_1, \dots, X_p)
        = P(\omega X_1, \dots, \omega X_p) 
        \quad .
    \end{equation*}
    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
    as polynomials,
    and in particular that
    the \kl{maximal monomials} of 
    $P$ are \kl{non-negative}.

    Remark that for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
    the polynomial $\restr{P}{\nu}$ continues to be represented
    by a \kl{$\Nat$-polyregular function}, namely
    $f$ partially applied to a word. As a consequence,
    the \kl{maximal monomials} of
    $\restr{P}{\nu}$ are also \kl{non-negative}, 
    and
    we have proven that $P \in \CorrectPoly$.
\end{proofof}

@@

\begin{proofof}[derivation-monomials:fact]
    Let $M \defined X_1^{k_1} \cdots X_n^{k_n}$,
    where $k_1, \dots, k_n > 0$ and $n > 0$.
    Let us compute
    \begin{align*}
        \Diff{K}{M} &= \prod_{i = 1}^n X_i^{k_i} - \prod_{i = 1}^n (X_i - K)^{k_i} \\
                    &= \sum_{p = 1}^n \frac{K}{X_p - K} \prod_{i = 1}^n (X_i - K)^{k_i} 
    \end{align*}
    Therefore, 
    the \kl{maximal monomials} of $\Diff{K}{M}$
    are precisely
    of the form $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$ for some $1 \leq p \leq i$,
    which \kl{strictly divides} $M$.
    Let $T$ be a \kl{monomial} that \kl{strictly divides} $M$,
    then, $T = \prod_{i = 1}^n X_i^{t_i}$, 
    for all $1 \leq i \leq n$, $t_i \leq k_i$,
    and there exists $1 \leq p \leq n$ such that$ t_j < k_j$.
    In particular, $T$ \kl{divides}
    $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$.
\end{proofof}

\begin{proofof}[derivation-simplifies:lemma]
    Let $P \in \Rel[\vec{X}]$ and
    $K \in \Nat$.
    Leveraging 
    \cref{derivation-monomials:fact,discrete-deriv-linear:fact,maxi-monomials-submodular:fact}:
    \begin{align*}
        \MaximalMonomials\left(\Diff{K}{P}\right)
        &=
        \MaximalMonomials\left(\sum_{M \in \Monomials(P)} \Diff{K}{M}\right) \\
        &\hoareleq
        \bigcup_{M \in \Monomials(P)} 
        \MaximalMonomials(\Diff{K}{M}) \\
        &\hoarele
        \MaximalMonomials(P)
        \quad .
        \qedhere
    \end{align*}
\end{proofof}

\begin{proofof}[derivation-stabilises-correct:lem]
    If $P$ is constant, then $P_2 = 0$, and $\Diff{K}{P_1} = 0$,
    hence for $K = 0$, $Q \in \CorrectPoly$.

@@

    Now, remark that the coefficient $\beta$ is obtained as follows:
    \begin{align*}
        \beta &= \sum_{M \in \mathcal{D}_1} \restr{M / T_1}{\nu}
               + \sum_{M \in \mathcal{D}_2} \restr{M / T_1}{\nu} \\
              &\geq \sum_{M \in \mathcal{D}_1} \restr{M / T_1}{\nu}
              - \sum_{M \in \mathcal{D}_2} \left|\restr{M / T_1}{\nu}\right|
              \\
              &\geq
              \sum_{M \in \mathcal{D}_1}
              \left[
              \restr{M / T_1}{\nu}
              -
              \sum_{S \in \mathcal{D}_2^M}
              \left|\restr{S/T_1}{\nu}\right|
              \right]
              \\
              &\geq
              \sum_{M \in \mathcal{D}_1}
              \restr{\mathsf{unit}(M)/T_1}{\nu}
              \left[
              K
              - \alpha D
              \right]
              \\
              &\geq 0 \quad . \qedhere
    \end{align*}
\end{proofof}

\begin{proofof}[derivation-translation:lem]

@@

    For the effective decision procedure,
    given $P \in \Rel[\vec{X}]$,
    one applies \cref{derivation-translation:lem}
    to obtain a bound $K$, and then 
    computes $\translate{K}(\restr{P}{\nu})$ for all 
    partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots, K}$.
    Then $P \in \CorrectPoly$ if and only if all those polynomials
    belong to $\Nat[\vec{X}]$, which is decidable.
\end{proofof}

\begin{proofof}[lem:correct-covered-2]
    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
    by considering the empty valuation $\nu \colon \set{X,Y} \topartial \Nat$.
    For the converse inclusion, let us consider $P(X,Y)$
    that is \kl{non-negative}, such that the \kl{maximal monomials}
    are  \kl{non-negative}.

    If we fix none of the variables, then the \kl{maximal monomials}
    are \kl{non-negative} by assumption. If we fix one of the
    variables, we can assume without loss of generality that we 
    fix $X = k$ for some $k \in \Nat$.
    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
    and therefore must have a positive leading coefficient
    (which is the unique \kl{maximal monomial} in this case)
    or be constant equal to 0. In both cases, the \kl{maximal monomials}
    have positive coefficient.
    The same reasoning applies \emph{a fortiori} in the case where
    we fix the two indeterminates, leading to a constant polynomial.
\end{proofof}

### 2024-03-29 parts/conclusion.tex
appendix + proofrefs

In the commutative case, the conjectured semantic characterizations of $\NSF$
and $\NPoly$ have been proven to hold. Furthermore, the existence of a
canonical transducer computing functions in $\NPoly$ opens the door to several
questions with respect to aperiodicity. We can prove that for functions in
$\NPoly[0]$, aperiodicity is equivalent to the absence of counters in the
\kl{residual transducer} of the function (see
\cref{aperiodic-iff-residual:lem}), however this result does not generalize to
higher growth rates, as witnessed by \cref{non-aperiodic-residual-transd:ex}.
In the example, no \kl{$\NPoly[0]$-transducer} with the minimal number
of states computing the function is \kl{counter-free}, which suggests that a
more sophisticated object is needed to observe aperiodicity.

\begin{lemma}
    \label{aperiodic-iff-residual:lem}
    Let $f \in \NPoly[0]$, the following are equivalent:
    \begin{enumerate}
        \item $f \in \NSF$,
        \item $f \in \ZSF$,
        \item $f$ is \kl{ultimately polynomial},
        \item The \kl{$0$-residual transducer} of $f$ is \kl{counter-free}.
    \end{enumerate}
\end{lemma}

\begin{example}
    \label{non-aperiodic-residual-transd:ex}
    Let us define
    $\BadExKo(\varepsilon) = 1$,
    $\BadExKo(a) = 0$,
    $\BadExKo(a^2) = 1$,
    and $\BadExKo(a^n) = n - 3$ for all $n \geq 3$.
    The \kl{$0$-residual transducer} of $\BadExKo$ has a \kl{counter} and two states.
    Furthermore,
    every \kl{$\NPoly[0]$-transducer} with two states contains a \kl{counter}.
\end{example}

\AP In the spirit of the characterization of $\NPoly$ based the
$(\resleq{f}{k})$ quasi-ordering (\cref{non-commutative-npoly:thm}), we
introduce the notion of \intro{aperiodic ordering} of $\Sigma^*$ as follows:
$(\Sigma^*, \leq)$ is \reintro[aperiodic ordering]{aperiodic} whenever for all
$u, w \in \Sigma^*$, there exists $N_0 \in \Nat$, such that the sequence
$\seqof{uw^n}{n \geq N_0}$ is increasing. This leads to the following
\cref{sf-no-periods-on-sequences:conj}, of which the easy implication is
already known to hold (\cref{sf-no-periods-on-sequences:lemma}), and requires
the introduction of the \emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of
$(\resleq{f}{k})$, defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(\app{f}{v} - \app{f}{u}) \in
\NSF[k-1]$.

\begin{lemma}
    \label{sf-no-periods-on-sequences:lemma}
    Let $k \in \Nat$, and $f \in \SF[k]$. Then,
    $(\Sigma^*, \resleqsf{f}{k})$ is an
    \kl[aperiodic ordering]{aperiodic} \kl{well-quasi-ordering}.
\end{lemma}

\begin{conjecture}
    \label{sf-no-periods-on-sequences:conj}
    For all $k \in \Nat$ and $f \colon \Sigma^* \to \Rel$,
    $f \in \NSF[k]$ if and only if $(\Sigma^*, \resleqsf{f}{k})$ is an \kl[aperiodic ordering]{aperiodic}
    \kl{well-quasi-ordering}.
\end{conjecture}

Let us conclude with a more general remark on the status of commutative input
in the study of polyregular functions. Notice that semantic characterizations
of subclasses of \kl{polyregular functions} (\kl{combinatorial}, \kl{ultimately
polynomial}, pumpable) arise from pre-composition with a \kl{commutative}
\kl{star-free polyregular function}, as formally stated in
\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.
Generalizing this idea, we conjecture that an effective encoding of $\ZPoly$ in
terms of commutative functions could be obtained, as stated in
\cref{z-poly-commutative-encoding:conj}, playing the same role as the
decomposition obtained in \cref{decompose-polynomial:lem} in the
\kl{commutative} case.

\begin{remark}
    \label{pre-compose-growth-commut:lemma}
    Let $f \in \ZRat$, and $k \in \Nat$. Then,
    $f \in \ZPoly[k]$ if and only if 
    for every \kl{commutative} \kl{star-free polyregular function} $h$
            of \kl{growth rate} $l \in \Nat$,
            $(f \circ h) \in \ZPoly[k+l-1]$.
\end{remark}

\begin{remark}
    \label{pre-compose-sf-commut:lemma}
    Let $f \in \ZPoly$. Then, $f \in \ZSF$,
    if and only if for every \kl{commutative} \kl{star-free polyregular function} $h$,
            $(f \circ h) \in \ZSF$.
\end{remark}

@@

In the specific subclass of \emph{polyblind functions}
\cite{LENP21,DOUE22}, parts of the conjecture holds,
because they essentially behave as commutative functions \cite[Theorem
6.12]{DOUE23}. Indeed, a suitable application of Parikh's Theorem
\cite{PARI66} guarantees the existence of the functions $\mathsf{enc}$
and $\mathsf{dec}$, although they may fail to be  \kl{polyregular functions}.
Although \cref{z-poly-commutative-encoding:remark} is not yet applicable in
practice, we consider this partial result as a strong indicator in favor of the
conjecture.

\begin{remark}
    \label{z-poly-commutative-encoding:remark}
    The \cref{z-poly-commutative-encoding:conj} holds for the
    subclass of polyblind functions when dropping the assumption that
    $\mathsf{enc}$ and $\mathsf{dec}$ are \kl{polyregular functions}
    (resp. \kl{star-free polyregular functions}).
\end{remark}

>>>>

Let us conclude with a more general remark on the status of commutative input
in the study of unary output polyregular functions. Notice that all semantic
characterizations of subclasses of $\ZPoly$ (\kl{combinatorial}, \kl{ultimately
polynomial}, pumpable) arise from pre-composition with a \kl{commutative}
\kl{star-free polyregular function}, as formally stated in
\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}. Following
this idea, we conjecture that $\ZPoly$ and $\NPoly$ can be described as
combinations of commutative functions, leading to
\cref{z-poly-commutative-encoding:conj}. This conjecture would play the role of
\cref{decompose-polynomial:lem} in the study of functions in $\ZPoly$.

\begin{lemma}
    \label{pre-compose-growth-commut:lemma}
    Let $f \in \ZRat$, and $k \in \Nat$. Then,
    $f \in \ZPoly[k]$ if and only if 
    for every \kl{commutative} \kl{star-free polyregular function} $h$
            of \kl{growth rate} $l \in \Nat$,
            $(f \circ h) \in \ZPoly[k+l-1]$.
\end{lemma}

\begin{lemma}
    \label{pre-compose-sf-commut:lemma}
    Let $f \in \ZPoly$. Then, $f \in \ZSF$,
    if and only if for every \kl{commutative} \kl{star-free polyregular function} $h$,
            $(f \circ h) \in \ZSF$.
\end{lemma}

@@

In the specific subclass of \emph{polyblind functions}
\cite{LENP21,DOUE22}, parts of the conjecture holds,
because they essentially behave as commutative functions \cite[Theorem
6.12]{DOUE23}. Indeed, a suitable application of Parikh's Theorem
\cite{PARI66} guarantees the existence of the functions $\mathsf{enc}$
and $\mathsf{dec}$, although they may fail to be  \kl{polyregular functions}.

\begin{lemma}
    \label{z-poly-commutative-encoding:remark}
    The \cref{z-poly-commutative-encoding:conj} holds for the
    subclass of polyblind functions when dropping the assumption that
    $\mathsf{enc}$ and $\mathsf{dec}$ are \kl{polyregular functions}
    (resp. \kl{star-free polyregular functions}).
\end{lemma}

### 2024-03-29 parts/correction.tex
appendix + proofrefs

The counter example provided by \cref{def:bad-polynomial} relies on the fact
that $\CoveredPoly$ is not stable under fixing indeterminates, while
\kl{$\Nat$-polyregular functions} are. In this section, we prove that closing
$\CorrectPoly$ under assignments of variable is enough to recover from
\cref{karh:thm}.

\AP We use the following notation to fix the value of some indeterminate, if
$P(X,Y)$ is a polynomial in $\Rel[X,Y]$, then $\intro*\restr{P(X,Y)}{X = 1}$ is
the polynomial $P(1,Y) \in \Rel[Y]$. More generally, if $\nu$ is a partial
function from $\vec{X}$ to $\Nat$, written $\nu \colon \vec{X} \topartial
\Nat$, the restriction $\restr{P(\vec{X})}{\nu}$ is the polynomial with
indeterminates $\vec{Y} \defined \vec{X} - \dom(\nu)$ obtained by fixing the
variables of the domain of $\nu$.

@@

\begin{lemma}
    \label{n-poly-combinatorics:lem}
    Let $f$ be a  \kl{$\Nat$-polyregular} function. 
    There exists a computable $\omega \in \Nat$
    such that for all $p \in \Nat$,
    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
    for all $u_1, \dots, u_p \in \Sigma^*$,
    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
    whose \kl{maximal monomials} are \kl{non-negative},
    and such that for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \alpha_0 \prod_{i = 1}^p u_i^{\omega X_i} \alpha_i
        \right)
        = P(X_1, \dots, X_p) \quad .
    \end{equation*}
\end{lemma}

\begin{corollary}
    \label{n-rat-correct:lem}
    Let $P \in \Rel[X_1, \dots, X_p]$ be \kl{represented}
    by a \kl{$\Nat$-rational series}. Then,
    $P \in \CorrectPoly$.
\end{corollary}

@@

\begin{fact}
    \label{derivation-monomials:fact}
    Let $K > 0$,
    and let $M,T$ be two \kl{monomials}, such that
    $T$ \kl{strictly divides} $M$.
    Then,
    every \kl{monomial} of $\Diff{K}{M}$ \kl{strictly divides} $M$,
    and 
    $T$ \kl{divides} some \kl{maximal monomial}
    of $\Diff{K}{M}$, which has a coefficient
    that is a multiple of $K$.
\end{fact}

\begin{lemma}
    \label{derivation-simplifies:lemma}
    For all $P \in \Rel[\vec{X}]$ that are non-constant,
    for all $K \in \Nat$,
    $\MaximalMonomials(\Diff{K}{P}) \hoarele
    \MaximalMonomials(P)$.
\end{lemma}

@@

\begin{lemma}
    \label{derivation-stabilises-correct:lem}
    Let $P \in \CorrectPoly$,
    $P_1$ be the sum of \kl{maximal monomials} of $P$,
    and $P_2 \defined P - P_1$ be the sum of
    non-maximal monomials of $P$.
    There exists a computable $K$,
    such that
    $Q \defined (\Diff{K}{P_1} + \translate{K}(P_2)) \in \CorrectPoly$.
\end{lemma}

@@

\begin{lemma}
    \label{derivation-translation:lem}
    Let $P \in \Rel[\vec{X}]$, 
    there exists a computable $K \in \Nat$
    such that the following are equivalent:
    \begin{enumerate}
        \item \label{d-t-correct:item} $P \in \CorrectPoly$,
        \item \label{d-t-transl:item}
            for 
            all partial functions $\nu \colon \vec{X} \topartial \Nat$,
            $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$,
        \item \label{d-t-transl-fin:item}
            for all partial functions
            $\nu \colon \vec{X} \topartial \set{0, \dots, K}$,
            $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$.
    \end{enumerate}
    Furthermore, the membership is decidable.
\end{lemma}

@@

Let us now state, for the sake of completeness, the equality of $\CorrectPoly$
and $\CoveredPoly$ in the case of two indeterminates. This is relevant because
it shows that examples given in \cite{KARH77} are actually correct, even though
based on an invalid result, and it may be the case for other works based on
\cite{KARH77}.

\begin{lemma}
    \label{lem:correct-covered-2}
    $\CorrectPoly[X,Y] = \CoveredPoly[X,Y]$.
\end{lemma}
\begin{proof}
    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
    by considering the empty valuation $\nu \colon \set{X,Y} \topartial \Nat$.
    For the converse inclusion, let us consider $P(X,Y)$
    that is \kl{non-negative}, such that the \kl{maximal monomials}
    are  \kl{non-negative}.

    If we fix none of the variables, then the \kl{maximal monomials}
    are \kl{non-negative} by assumption. If we fix one of the
    variables, we can assume without loss of generality that we 
    fix $X = k$ for some $k \in \Nat$.
    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
    and therefore must have a positive leading coefficient
    (which is the unique \kl{maximal monomial} in this case)
    or be constant equal to 0. In both cases, the \kl{maximal monomials}
    have positive coefficient.
    The same reasoning applies \emph{a fortiori} in the case where
    we fix the two indeterminates, leading to a constant polynomial.
\end{proof}

>>>>

\AP
The counter example provided by \cref{def:bad-polynomial} relies on the fact
that $\CoveredPoly$ is not stable under fixing indeterminates, while
\kl{$\Nat$-polyregular functions} are. In this section, we prove that closing
$\CorrectPoly$ under assignments of variable is enough to recover from
\cref{karh:thm}.
We use the following notation to fix the value of some indeterminate, if
$P(X,Y)$ is a polynomial in $\Rel[X,Y]$, then $\intro*\restr{P(X,Y)}{X = 1}$ is
the polynomial $P(1,Y) \in \Rel[Y]$. More generally, if $\nu$ is a partial
function from $\vec{X}$ to $\Nat$, written $\nu \colon \vec{X} \topartial
\Nat$, the restriction $\restr{P(\vec{X})}{\nu}$ is the polynomial with
indeterminates $\vec{Y} \defined \vec{X} - \dom(\nu)$ obtained by fixing the
variables of the domain of $\nu$.

@@

\begin{lemma}
    \label{n-poly-combinatorics:lem}
    Let $f$ be an \kl{$\Nat$-polyregular} function. 
    There exists a computable $\omega \in \Nat$
    such that for all $p \in \Nat$,
    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
    for all $u_1, \dots, u_p \in \Sigma^*$,
    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
    whose \kl{maximal monomials} are \kl{non-negative},
    and such that for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \alpha_0 \prod_{i = 1}^p u_i^{\omega X_i} \alpha_i
        \right)
        = P(X_1, \dots, X_p) \quad .
    \end{equation*}
    \proofref{n-poly-combinatorics:lem}
\end{lemma}

\begin{corollary}
    \label{n-rat-correct:lem}
    Let $P \in \Rel[X_1, \dots, X_p]$ be an \kl{$\Nat$-rational polynomial}.
    Then,
    $P \in \CorrectPoly$.
    \proofref{n-rat-correct:lem}
\end{corollary}

@@

\begin{fact}
    \label{derivation-monomials:fact}
    Let $K > 0$,
    and let $M,T$ be two \kl{monomials}, such that
    $T$ \kl{strictly divides} $M$.
    Then,
    every \kl{monomial} of $\Diff{K}{M}$ \kl{strictly divides} $M$,
    and 
    $T$ \kl{divides} some \kl{maximal monomial}
    of $\Diff{K}{M}$, which has a coefficient
    that is a multiple of $K$.
    \proofref{derivation-monomials:fact}
\end{fact}

\begin{lemma}
    \label{derivation-simplifies:lemma}
    For all $P \in \Rel[\vec{X}]$ that are non-constant,
    for all $K \in \Nat$,
    $\MaximalMonomials(\Diff{K}{P}) \hoarele
    \MaximalMonomials(P)$.
    \proofref{derivation-simplifies:lemma}
\end{lemma}

@@

\begin{lemma}
    \label{derivation-stabilises-correct:lem}
    Let $P \in \CorrectPoly$,
    $P_1$ be the sum of \kl{maximal monomials} of $P$,
    and $P_2 \defined P - P_1$ be the sum of
    non-maximal monomials of $P$.
    There exists a computable $K$,
    such that
    $Q \defined (\Diff{K}{P_1} + \translate{K}(P_2)) \in \CorrectPoly$.
    \proofref{derivation-stabilises-correct:lem}
\end{lemma}

@@

\begin{lemma}
    \label{derivation-translation:lem}
    Let $P \in \Rel[\vec{X}]$, 
    there exists a computable $K \in \Nat$
    such that the following are equivalent:
    \begin{enumerate}
        \item \label{d-t-correct:item} $P \in \CorrectPoly$,
        \item \label{d-t-transl:item}
            for 
            all partial functions $\nu \colon \vec{X} \topartial \Nat$,
            $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$,
        \item \label{d-t-transl-fin:item}
            for all partial functions
            $\nu \colon \vec{X} \topartial \set{0, \dots, K}$,
            $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$.
    \end{enumerate}
    Furthermore, the membership is decidable.
    \proofref{derivation-translation:lem}
\end{lemma}

@@

Let us now state, for the sake of completeness, the equality of $\CorrectPoly$
and $\CoveredPoly$ in the case of two indeterminates. This is relevant because
it shows that examples given in \cite{KARH77} are actually correct, even though
based on an invalid result, and it may be the case for other works based on
\cite{KARH77}. The proof is based the fact that for (non-negative) univariate
it is enough that the dominant coefficient is non-negative to be an
\kl{$\Nat$-rational polynomial}.

\begin{lemma}
    \label{lem:correct-covered-2}
    $\CorrectPoly[X,Y] = \CoveredPoly[X,Y]$.
    \proofref{lem:correct-covered-2}
\end{lemma}
